% ============================================================
%  SECTION 5: EMPIRICAL STRATEGY
% ============================================================

\section{Empirical Strategy}
\label{sec:empirical_strategy}

We deploy two empirical designs. The first is an adoption
difference-in-differences that identifies the effect of Tennessee HOPE
on college composition. The second is a shock-interaction design that
tests whether merit aid buffers low-income access during local economic
downturns, exploiting county-level income variation described in
Section~\ref{sec:data_bea}. Together, the two designs identify
different parameters---the level shift from program adoption and the
differential response to income shocks---using independent sources of
variation.

\subsection{Adoption DiD: Primary Estimator}

Treatment timing is sharp: all Tennessee public colleges are treated
simultaneously at $t^* = 1987$, the first birth cohort whose college
choice set was formed under the fully known HOPE regime (see
Section~\ref{sec:background_timeline} for the decision-year
justification). We designate 1986 ($k = -1$) as the omitted reference
period, absorbing any partial treatment of that transition cohort into
the baseline. Robustness to $t^* \in \{1985, 1986, 1987\}$ is
reported in Section~\ref{sec:robustness}; the static estimate is
insensitive to this choice.

Our primary estimator is the two-way fixed effects (TWFE) event study.
Because treatment timing is sharp---all Tennessee public colleges are
treated simultaneously---recent heterogeneity-robust estimators
designed for staggered adoption
\parencite{callaway2021difference, sun2021estimating} reduce to the
standard TWFE specification in our setting
\parencite{roth2026interpreting}. We therefore report TWFE as
the primary estimator and reserve the
\textcite{callaway2021difference} estimator for the pooled multi-state
specification in Section~\ref{sec:discussion}, where staggered
adoption timing makes it the appropriate choice. The TWFE
specification is:

\begin{equation}
y_{ct} = \alpha_c + \lambda_t + \sum_{k=-5}^{4} \beta_k \cdot
\mathbf{1}[t - t^* = k] \cdot \text{Treated}_c + \gamma \cdot
\overline{\log(\text{pcpi})}_{c} + \varepsilon_{ct}
\label{eq:twfe}
\end{equation}

\noindent where $y_{ct}$ is the outcome for college $c$ in birth cohort
$t$, $\alpha_c$ are college fixed effects, $\lambda_t$ are cohort fixed
effects, $\text{Treated}_c$ indicates Tennessee public colleges,
$\overline{\log(\text{pcpi})}_{c}$ is the pre-treatment county average
of log per capita personal income (BEA, 1997--2003 decision years),
and $k = -1$ (cohort 1986) is the omitted reference period. The sample
excludes the 1980 birth cohort, a single-cohort outlier that drives
pre-trend divergence at the longest pre-treatment horizon. Cohort 1981
($k = -6$) remains in the estimation sample---contributing to the
precision of the college and cohort fixed effects---but is not assigned
a separate event-time indicator; its effect is subsumed into the
intercept.

\subsection{Preferred Specification}

The preferred specification for the adoption DiD uses:
\begin{enumerate}
  \item \textbf{National control group}: public colleges in 26
    non-merit-program states, providing 27 state-level clusters
    (26 control $+$ Tennessee) for cluster-robust inference.
  \item \textbf{Treatment timing $t^* = 1987$}: the first
    decision-treated cohort
    (Section~\ref{sec:background_timeline}).
  \item \textbf{Drop 1980 birth cohort}: Tennessee's 1980-born cohort
    exhibits anomalously low $Q_1$ share that normalizes by 1982,
    consistent with a data artifact. This single cohort is the sole
    driver of the $k = -6$ pre-trend divergence in prior
    specifications.
  \item \textbf{BEA county income control}: pre-treatment county
    average log per capita personal income (BEA, 1997--2003 decision
    years) absorbs cross-county differences in baseline economic
    conditions. Because the average is computed entirely from
    pre-treatment cohorts, it is predetermined and not affected by
    HOPE. The decomposition framework
    (Section~\ref{sec:theory_decomposition}) identifies local
    economic conditions as the channel through which income effects
    operate, motivating the control independently of the sensitivity
    analysis.
\end{enumerate}

Each choice is independently motivated and precedes the sensitivity
analysis. Results without the BEA control are nearly identical
(Section~\ref{sec:robustness}), confirming that the baseline county
income level is approximately collinear with the college fixed
effects.

For robustness, we also report results under the bordering-state
control group (Alabama, Virginia, North Carolina---the only non-merit
states adjacent to Tennessee), which has cleaner pre-trends but only
four state clusters, as well as results under alternative treatment
timing.

\subsection{Identification}

Identification requires that Tennessee public colleges would have
followed parallel trends in socioeconomic composition to control
colleges absent the HOPE Scholarship. The most plausible threat is
that Tennessee's public sector was on a different compositional
trajectory before HOPE---driven, for example, by demographic shifts,
state fiscal policy, or differential enrollment growth---and that
this trajectory, not the scholarship, produces the post-treatment
pattern.

Under the preferred specification, pre-treatment event-study
coefficients at $k = -4$, $-3$, and $-2$ (cohorts 1983--1985) are
individually small and statistically insignificant for $Q_1$, with
p-values exceeding 0.44. A joint test of the three pre-treatment
coefficients fails to reject the null of zero pre-trend. Given the
limited number of pre-treatment periods, these tests have modest
power against slow-moving violations; we address this directly with
the \textcite{rambachan2023more} sensitivity analysis below.

The directional pattern of the pre-treatment coefficients provides
additional reassurance. The coefficient at $k = -3$ is the largest
in absolute value and is negative: Tennessee's $Q_1$ share was
\emph{falling} relative to controls before HOPE. Post-treatment
effects are uniformly positive. A smooth confounding trend that
generates positive post-treatment effects would require the pre-trend
to be positive and growing---the opposite of what we observe. The
sign reversal is inconsistent with a continuous confound driving our
results, though it does not rule out a discrete break coinciding
exactly with HOPE's introduction.

\subsection{Shock-Interaction Design}
\label{sec:empirical_shock}

The decomposition framework (Section~\ref{sec:theory_decomposition})
predicts that tuition-scaling scholarships buffer college composition
during income downturns (P8), while fixed-dollar programs do not (P9)
and may even worsen composition at four-year colleges (P10). To test
these predictions, we estimate whether composition responds
differently to local economic shocks in states with merit-aid
programs:

\begin{equation}
y_{ct} = \alpha_c + \lambda_t + \beta_1 \cdot \text{Shock}_{c,t+17}
+ \beta_2 \cdot (\text{Shock}_{c,t+17} \times
\text{MeritDesign}_{s(c)}) + \varepsilon_{ct}
\label{eq:shock}
\end{equation}

\noindent where $\text{Shock}_{c,t+17}$ is the one-year log change in
BEA county per capita personal income at decision year, and
$\text{MeritDesign}_{s(c)}$ characterizes the merit-aid program in the
state of college $c$. We estimate this for three definitions of
$\text{MeritDesign}$: (i) a binary indicator for any strong merit
program, (ii) separate indicators for tuition-scaling and fixed-dollar
programs, and (iii) individual state interactions.

The coefficient $\beta_1$ captures the baseline relationship between
income shocks and composition in non-merit states. The coefficient
$\beta_2$ captures the \emph{differential} response in merit-aid
states---the insurance or reallocation channel.

The sample for the shock-interaction design includes all states (not
just Tennessee and controls), with public and private nonprofit colleges
(excluding for-profits, which are not covered by merit programs). This
is a cross-sectional comparison: we do not require a pre-program
period, so states where the merit program was already active throughout
the Chetty panel (Georgia since 1993, Florida since 1997) are fully
usable. Standard errors are clustered at the college level.

The identifying assumption is that income shocks would shift
composition equally in merit and non-merit states absent the programs.
College and cohort fixed effects absorb level differences and common
time trends; we require only that \emph{shock responsiveness}---not
levels---is similar across program-design groups. This assumption is
weaker than the adoption DiD's parallel trends requirement. The most
plausible threat is selection into program adoption: states that chose
tuition-scaling designs may differ in pre-existing aid generosity or
labor-market structure in ways that independently affect shock
responsiveness.

\subsection{Inference}

For the adoption DiD, standard errors are clustered at the state level,
the level at which treatment is assigned. The national specification
has 27 clusters (26 control states $+$ Tennessee). This is above the
conventional threshold for reliable cluster-robust inference, though
the extreme imbalance (one treated cluster vs.\ 26 controls) warrants
caution \parencite{mackinnon2017wild}. For the bordering-state
specification (4 clusters), conventional cluster-robust inference is
unreliable; \textcite{lee2026simple} show that collapsing the panel
into a cross-sectional regression after unit-specific demeaning or
detrending permits exact $t$-based inference and randomization
inference even with a single treated unit. We supplement with
Fisher-style permutation tests that assign placebo treatment to each
state in turn (Section~\ref{sec:robustness_permutation}).

For the shock-interaction design, standard errors are clustered at the
college level ($\sim$500 clusters for non-merit states, $\sim$1,200
total), providing ample degrees of freedom.

\subsection{Sensitivity Analysis}

We implement the honest confidence intervals of
\textcite{rambachan2023more}, which relax the parallel trends assumption
by allowing for bounded violations. We report results under two
restriction classes:

\begin{itemize}
\item \textbf{Relative magnitude} ($\Delta_\text{RM}$): post-treatment
  violations of parallel trends are bounded by a multiple $\bar{M}$ of
  the maximum pre-treatment violation. At $\bar{M} = 1$, post-treatment
  violations can be as large as the worst observed pre-treatment
  violation and the confidence interval must still exclude zero.

\item \textbf{Smoothness} ($\Delta_\text{SD}$): the rate of change
  of trend violations is bounded by $M$. This restriction is
  appropriate when the concern is a gradually accelerating confound
  rather than a discrete level shift.
\end{itemize}

\noindent The preferred specification produces substantial robustness
under both restrictions; we report the specific breakdown values and
the comparison across specifications in
Section~\ref{sec:robustness}.

\subsection{Synthetic Control}

As a complementary strategy, we implement the synthetic control method
\parencite{abadie2010synthetic}, constructing a data-driven
counterfactual for Tennessee from weighted combinations of donor
states. The donor pool consists of 22 non-merit states, matching on
pre-treatment bottom-quintile share levels (birth cohorts 1981--1985,
consistent with the preferred specification's sample). This approach
avoids imposing equal weights across all control states and directly
addresses concerns about the suitability of any particular comparison
group. We report the synthetic control trends in
Section~\ref{sec:results} and placebo-in-space inference in
Section~\ref{sec:robustness_synth}.
