% ============================================================
%  FRONT MATTER: Title, Authors, Abstract
% ============================================================

% ---- Title ----
\title{\LARGE\bfseries Merit Aid, Sorting, and Institutional Access:%
  \\[0.3em] Evidence from Tennessee's HOPE Scholarship%
  \thanks{We are grateful to
  \href{mailto:dg3342@columbia.edu}{Daniel Garcia},
  \href{mailto:fam2175@columbia.edu}{Francine Montecinos Puente}, and
  \href{https://sites.google.com/view/seyhan-erden/home}{Seyhan Erden}
  (Columbia University);
  \href{https://www.hridaykarnani.com/}{Hriday Karnani}
  (Harvard University); and
  \href{mailto:as7596@columbia.edu}{Andrea Scalenghe}
  (MIT Sloan)
  for helpful discussions and comments. We also thank the Federal Reserve
  Bank of Cleveland and the Peer Review Board of the Economic Scholars
  Program. The views expressed herein are those of the authors and do not
  necessarily reflect the views of the Federal Reserve Bank of New York
  or Columbia University.}}

\author{%
  \vspace{-0.2em}\\
  {\large
  \href{mailto:bok2106@columbia.edu}{Balint Kidd} \quad
  \href{mailto:trs2169@columbia.edu}{Tommy Soltanian} \quad
  \href{mailto:tyler.sotomayor@columbia.edu}{Tyler Sotomayor} \quad
  \href{mailto:gu2145@columbia.edu}{Gabriel Uceda-Sosa}} \\[0.7em]
  {\large Columbia University} \\[0.7em]
  {\normalsize April 2026} \\[1.2em]
  {\small\color{blue}\href{https://www.tylersotomayor.com/merit}{Link to Most Current Version}}
}

\date{\vspace{-3em}}

% ============================================================
\begin{document}
% ============================================================

\workingpaperheader

\begingroup
\onehalfspacing
\maketitle

\vspace{1em}

\begin{abstract}
\noindent
Merit-based scholarship programs are widely viewed as regressive because
eligibility correlates with family income. We show that this conclusion
conflates individual attendance effects with institutional composition
effects---distinct estimands that can move in opposite directions when
students sort across multiple sectors. Using tax-linked data from
\textcite{chetty2020mobility}, we find that Tennessee's HOPE Scholarship
reshapes the full parental income distribution at public colleges: all
five income quintiles shift monotonically, with bottom-quintile share
rising and top-quintile share falling. Total enrollment is unchanged;
the shift reflects three-destination sorting in which lower-income
students enter from non-attendance while upper-income students exit to
out-of-state alternatives. A permutation test ($p = 0.037$ for parental
income rank, $p = 0.074$ for bottom-quintile share) and synthetic
control analysis confirm Tennessee as an outlier. Results are robust
across four heterogeneity-corrected estimators. Massachusetts' Adams
Scholarship independently produces a progressive shift of similar
magnitude under similar design parameters, while South Dakota's
high-threshold program produces the opposite pattern---as the model
predicts. A three-destination sorting framework reconciles these
progressive composition effects with the regressive attendance findings
of \textcite{dynarski2000hope} and identifies program design parameters
that determine sorting direction.
\end{abstract}

\vspace{1em}

\noindent \textbf{JEL Codes:} I23, I24, I28, H75, J24, C23

\vspace{0.25em}

\noindent \textbf{Keywords:} merit aid, college composition, sorting, HOPE Scholarship, difference-in-differences

\endgroup

\doublespacing
\newpage

\tableofcontents
\newpage
