% ============================================================
%  SECTION 3: THEORETICAL FRAMEWORK
% ============================================================

\section{Theoretical Framework}
\label{sec:theory}

We develop a model of college choice that generates testable
predictions about how merit aid reshapes the income composition of
public institutions and whether that reshaping is stable under
economic fluctuations. The key departure from standard two-sector
models is that we allow three destinations: in-state public college,
in-state private college, and an outside option (out-of-state
enrollment or non-attendance). We then introduce income shocks to decompose the scholarship's
compositional effect into a price channel and an income-sensitivity
channel, analogous in spirit to the Slutsky decomposition.

\subsection{Setup}

Consider a population of families indexed by income $y_i$ and student
ability $a_i$, where $y_i$ and $a_i$ are positively correlated. Each
family chooses among three options: a public college ($C_P$), a private
college ($C_R$), or no college ($0$). The indirect utility of family
$i$ from option $j \in \{P, R, 0\}$ is
\begin{equation}
  V_{ij} = q_j - \frac{C_j}{y_i} + \varepsilon_{ij},
\end{equation}
where $q_j$ is the quality of institution $j$, $C_j$ is the sticker
price, and $\varepsilon_{ij}$ captures idiosyncratic preferences (we
place no distributional restriction on $\varepsilon_{ij}$; the
propositions below use only the sign of expected marginal utilities).
We normalize $q_0 = C_0 = 0$ and assume:
\begin{enumerate}
  \item $q_R > q_P > 0$ (private colleges are higher quality),
  \item $C_R > C_P > 0$ (private colleges are more expensive),
  \item $y_i > 0$ for all families,
  \item $\text{Cov}(a_i, y_i) > 0$ (ability and income are positively associated).
\end{enumerate}

\subsection{Merit Scholarship and Sorting Boundaries}

A state merit scholarship of amount $S > 0$ reduces the effective price
of the public college for students with ability above a threshold
$\tau$:
\begin{equation}
  C_P^{\text{eff}}(a_i) =
  \begin{cases}
    C_P - S & \text{if } a_i \geq \tau, \\
    C_P     & \text{if } a_i < \tau.
  \end{cases}
\end{equation}

Without the scholarship, families sort across options by income. Define
the no-college/public boundary as $\underline{y}$, where $V_{iP} =
V_{i0}$, and the public/private boundary as $\overline{y}$, where
$V_{iP} = V_{iR}$:
\begin{equation}
  \underline{y} = \frac{C_P}{q_P}, \qquad
  \overline{y} = \frac{C_R - C_P}{q_R - q_P}.
\end{equation}
Families with $y_i < \underline{y}$ do not attend college; those with
$\underline{y} \leq y_i < \overline{y}$ attend public; those with
$y_i \geq \overline{y}$ attend private.

The scholarship shifts both boundaries for eligible students.

\begin{proposition}[Access margin]
\label{prop:access}
The scholarship lowers the no-college/public boundary:
$\underline{y}^S = (C_P - S)/q_P < \underline{y}$.
\end{proposition}

\begin{proof}
Since $S > 0$, we have $C_P - S < C_P$. Dividing by $q_P > 0$
preserves the inequality.
\end{proof}

\begin{proposition}[Substitution margin]
\label{prop:substitution}
The scholarship raises the public/private boundary:
$\overline{y}^S = \overline{y} + S/(q_R - q_P) > \overline{y}$.
\end{proposition}

\begin{proof}
Since $S > 0$ and $q_R - q_P > 0$, the term $S/(q_R - q_P) > 0$.
\end{proof}

\begin{proposition}[Progressivity of utility gain]
\label{prop:progressivity}
For an eligible student choosing public college, the utility gain from
the scholarship is $\Delta V_i = S / y_i$, which is strictly
decreasing in income: $\partial \Delta V_i / \partial y_i = -S/y_i^2 < 0$.
\end{proposition}

Together, the access and substitution margins predict that the public
sector's income composition shifts downward, with the largest gains at
the bottom of the distribution
(Proposition~\ref{prop:progressivity}).
Figure~\ref{fig:sorting_model} illustrates the sorting mechanism.

\input{tikz/fig_sorting_model}

\subsection{Testable Predictions}

The model generates seven predictions about the compositional effect
of merit aid:

\begin{description}
\item[P1: Bottom-quintile share rises at public colleges.] The
  scholarship reduces the effective price of public college. Because
  $S / y_i$ is largest for low-income families, the marginal entrants
  drawn from below $\underline{y}$ are disproportionately low-income.
  We test P1 in Section~\ref{sec:results_main}.

\item[P2: Monotonic quintile gradient.] The price reduction $S / y_i$
  is monotonically decreasing in income. Q1 families experience the
  largest proportional price reduction; Q5 the smallest. The
  composition shift should be monotonically graded across quintiles.
  We test P2 in Section~\ref{sec:results_quintile}.

\item[P3: Composition change reflects sorting, not enrollment
  expansion.] Total enrollment at public colleges is unchanged if
  capacity constraints bind; the share shifts are driven by changes in
  who enrolls, not in how many enroll.
  We test P3 in Section~\ref{sec:results_mechanisms}.

\item[P4: No mirror image at private colleges.] If displaced
  top-quintile students exit to out-of-state options rather than
  shifting to private colleges, private institutions should show no
  significant composition change.
  We test P4 in Section~\ref{sec:results_mechanisms}.

\item[P5: Heterogeneous channels by tier.] At two-year colleges, where
  the no-college margin is active, the scholarship expands access
  (pulls in new students from below $\underline{y}$). At four-year
  universities, where the public/private margin is active, the
  scholarship induces sorting---a form of the ability stratification
  documented by \textcite{cornwell2006sorting}.
  We test P5 in Section~\ref{sec:results_mechanisms}.

\item[P6: Spatial heterogeneity.] Institutions near state borders,
  where out-of-state alternatives are accessible, should show larger
  composition shifts because the outside option is more salient and the
  three-destination mechanism is most active.
  We test P6 in Section~\ref{sec:spatial}.

\item[P7: Cross-state variation.] Programs with no Pell offset and
  large public-private subsidy differentials produce progressive
  sorting. Programs with Pell offsets or comparable private awards
  produce regressive sorting or no composition change.
  We test P7 in Section~\ref{sec:discussion}.
\end{description}

\subsection{Decomposition Under Income Shocks}
\label{sec:theory_decomposition}

Predictions P1--P7 describe the \emph{static} compositional effect of
introducing a merit scholarship. We now ask whether that compositional
shift is \emph{stable} when families experience local economic
fluctuations. The answer depends on the scholarship's award structure.

Consider a negative income shock $\Delta y < 0$ affecting families in
a local economy. The shock changes family income from $y_i$ to
$y_i + \Delta y$ but does not change the scholarship amount, tuition,
or institutional quality. The total effect of the scholarship under an
income shock decomposes into two components.

\paragraph{Price effect.}
Holding income constant, the scholarship changes the relative price of
public college. The utility gain for an eligible student choosing
public is $\Delta V_i = S/y_i$. This price-reduction effect generates
predictions P1--P7. The adoption difference-in-differences identifies
it. It does not vary with local economic conditions.

\paragraph{Income-sensitivity effect.}
Holding the scholarship constant, the income shock changes the
family's ability to pay the remaining cost. For a fixed-dollar
scholarship $S$, the cost burden after the shock is
$(C_P - S)/(y_i + \Delta y)$, which increases when $\Delta y < 0$.
The income shock reasserts the affordability constraint on any costs
the scholarship does not cover.

The degree to which the scholarship \emph{compensates} the income
effect depends on the award structure:

\begin{proposition}[Income-effect compensation]
\label{prop:compensation}
Let $S^F$ be a fixed-dollar award and $S^\phi = \phi C_P$ be a
percentage-based award covering fraction $\phi$ of tuition, with
$S^F = \phi C_P$ so the two have identical dollar value at the time
of introduction. Under a percentage-based award, the income effect
on tuition costs is reduced by factor $(1 - \phi)$ relative to the
no-scholarship case. Under a fixed-dollar award, the reduction is
$1 - S^F/C_P$, which equals $(1 - \phi)$ initially but erodes as
tuition rises: if $C_P' > C_P$, then $1 - S^F/C_P' > 1 - \phi$.
\end{proposition}

\noindent When $\phi = 1$ (full tuition coverage), tuition costs
vanish from the income-sensitivity channel entirely: the family's
college choice on the tuition dimension is fully insulated from income
fluctuations.

\begin{proposition}[Tier reallocation under fixed-dollar awards]
\label{prop:tier}
Consider a student eligible for a fixed-dollar scholarship $S$,
choosing between a four-year college (cost $C_4$, quality $q_4$) and
a two-year college (cost $C_2 < C_4$, quality $q_2 < q_4$). The
scholarship reduces costs at both tiers by the same dollar amount $S$,
so the relative price of the four-year option is unchanged:
\begin{equation}
  V_{i4}^S - V_{i2}^S
  = (q_4 - q_2) - \frac{C_4 - C_2}{y_i}.
\end{equation}
For the marginal four-year student (where $V_{i4}^S \approx V_{i2}^S$),
an income shock $\Delta y < 0$ increases $(C_4 - C_2)/(y_i + \Delta y)$,
tipping the balance toward the two-year option. Provided
$y_i + \Delta y > (C_2 - S)/q_2$ (the student remains above the
no-college boundary), the scholarship keeps the student in the system
but does not prevent tier reallocation. For sufficiently large shocks,
exit from the system is also possible.
\end{proposition}

\noindent Proposition~\ref{prop:tier} explains why fixed-dollar
programs can generate progressive sorting on average (P1--P2) while
showing no aggregate buffering during downturns: the scholarship
creates a new class of marginal four-year students---low-income
families who attend four-year colleges only because the scholarship
makes it affordable---who are particularly sensitive to income shocks
and downshift to two-year colleges when family budgets tighten.
\textcite{bruce2014jackpot} document a consistent pattern for
Tennessee HOPE: marginal students near the eligibility threshold
shift from two-year to four-year institutions when HOPE is available.

Figure~\ref{fig:decomposition} illustrates the decomposition
spatially. In steady state, both Tennessee (fixed-dollar) and Florida
(percentage-based) produce the same progressive boundary shifts. Under
an income shock, Tennessee's boundaries retreat substantially---marginal
four-year students downshift to two-year---while Florida's boundaries
barely move.

\input{tikz/fig_decomposition}

These propositions generate four additional testable predictions:

\begin{description}
\item[P8: Tuition-scaling programs buffer composition during
  downturns.] Programs where the award covers a fixed share of tuition
  ($\phi > 0$) compensate the income effect, stabilizing the
  compositional shift achieved through the substitution effect.
  We test P8 in Section~\ref{sec:results_insurance}.

\item[P9: Fixed-dollar programs show no aggregate buffering.] Programs
  where the award is a fixed dollar amount leave the income effect
  uncompensated, producing zero net buffering at the aggregate level.
  We test P9 in Section~\ref{sec:results_insurance}.

\item[P10: Fixed-dollar programs show four-year anti-buffering.]
  At four-year colleges specifically, fixed-dollar programs should show
  composition \emph{worsening} during downturns---the marginal
  students created by the scholarship downshift to two-year colleges.
  We test P10 in Section~\ref{sec:results_tier_reallocation}.

\item[P11: Q2 is the most elastic quintile.] The lower-middle class
  (second income quintile) sits at the intersection of the model's two
  margins---just above the no-college boundary (where Pell alone may
  not suffice) and just below the two-year/four-year boundary (where
  the scholarship tips the choice). Q2 students should be the most
  sensitive to both the scholarship and to income shocks.
  We test P11 in Section~\ref{sec:results_tier_reallocation}.
\end{description}

\subsection{Reconciling with Dynarski (2000)}

\textcite{dynarski2000hope} finds that Georgia's HOPE Scholarship
widened the attendance gap between higher- and lower-income students.
Our framework shows that this aggregate finding is consistent with a
progressive composition shift at public institutions. Because
eligibility is correlated with income, merit aid increases the
\emph{total} number of higher-income students attending college more
than the total number of lower-income students. But the
\emph{destination} of these additional higher-income students matters.
If they enroll at private or out-of-state institutions---or if they
would have attended college anyway---the composition of public colleges
can still shift downward.

The reconciliation rests on distinguishing between two quantities: the
aggregate attendance gap (which can widen) and the institutional
composition of the public sector (which can simultaneously become more
progressive). Both are consistent with three-destination sorting.

This distinction is specific to the multi-sector setting. In a
two-sector model---attend college or not---the aggregate attendance
effect fully determines composition: if merit aid increases enrollment
disproportionately among higher-income students, those students
necessarily appear at colleges, and composition becomes more affluent.

With three or more destinations, the link breaks. Higher-income
students may attend college at higher rates but sort into private or
out-of-state institutions. The public sector's composition then depends
on who remains---not on aggregate enrollment. Individual enrollment
responses are therefore insufficient to predict institutional
composition; the full destination distribution is required.

\subsection{Georgia as a Regressive Prediction}
\label{sec:theory_georgia}

Georgia's HOPE included a Pell offset: scholarship dollars were reduced
dollar-for-dollar by Pell Grant receipts (Figure~\ref{fig:sorting_model}, Panel~C). Because Pell Grants target Q1
families, the effective scholarship for the poorest students was
\begin{equation}
  S_{\text{eff}}^{Q1} = \max(S - \text{Pell},\; 0) \approx 0.
\end{equation}
The model predicts no composition shift for Q1 (the price reduction is
zero), while Q3--Q5 families receive the full $S$. The predicted effect
is regressive: public colleges gain middle- and upper-income students
without gaining low-income students. \textcite{dynarski2000hope}
documents the consequence: Black students, who were disproportionately
Pell-eligible, showed no attendance gain, while white students gained
12.3 percentage points.

Georgia's HOPE also provided comparable awards to private colleges,
muting the public-private subsidy differential.
\textcite{cornwell2006enrollment} confirm the pattern: four-year private
enrollment in Georgia \emph{rose} by 13.2\% after HOPE, and two-thirds
of the four-year enrollment gain reflected reduced out-of-state
migration rather than new college-goers.

Tennessee's HOPE has no Pell offset, no income cap, and a large
public-private subsidy differential (full tuition at publics vs.\
\$3,000--\$4,000 at privates). Under these parameters, the model
predicts that the access margin and substitution margin both operate,
producing a progressive quintile gradient. The TN-GA contrast is
a pair of outcomes predicted by the same model under different
parameter values.

The decomposition framework adds a further distinction: Tennessee's
fixed-dollar award generates progressive sorting but leaves it
vulnerable to income shocks (uncompensated income effect), while
Florida's percentage-based award (Bright Futures Medallion tier,
covering 75\% of tuition) both generates progressive sorting and
stabilizes it (compensated income effect). Award structure predicts
not only the direction but the \emph{durability} of compositional
change.
